Reasonably Programmable Literal Notation

### 2018-06-22 icfp18/syntax-icfp18.tex
tighten up 2.2.1-2

\begin{figure}
\begin{lstlisting}[deletekeywords={spliced}]
  module Segment : { 
    type t = {start_pos: int, end_pos: int};
    let read_to : string => Lexing.lexbuf => Segment.t;
  };
  exception ExpansionError({msg: string, loc: option(Segment.t)});
  module ProtoExpr : { 
    type t = Parsetree.expression; /* from the OCaml compiler library */
    let spliced : Segment.t => Parsetree.core_type => t;
  };
  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
  notation $proto_expr at ProtoExpr.t = $re_expr;
\end{lstlisting}
\caption{A fragment of the \li{Relit} module signature relevant to expression TLM providers.}
\label{fig:relit-util}
\vspace{-4px}
\end{figure}
    % let match_spliced : t => option(Segment.t, ProtoTyp.t);

\subsubsection{\li{RegexLexer}}
\label{sec:regex-lexer}
% A fragment of the signature of the \li{Relit_helper} library, which provides various necessary definitions and utilities to TLM providers, appears in Fig.~\ref{fig:relit-util}. 
In most respects, the definition of the lexer in Fig.~\ref{fig:regex-lexer} is conventional. It reads various token patterns on Lines 6-23, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}. Lines 20-21 combine sequences of non-special characters (e.g. alphanumeric characters) into a single string token. Lines 22-23 implement backslash-prefixed escape sequences for characters that have special meaning in this notation for regexes, emitting a string token in each instance.\footnote{A better approach would be to define a second lexing rule that greedily combines sequences consisting of non-special characters and escape characters into a single \li{STR} token, but for the sake of exposition, we stick to this simpler approach.}

Lines 24-31 of the lexer, which recognize the notation for regex splicing \li{$(}$e$\li{)}, and string splicing, \li{$$(}$e$\li{)}, are more unusual. Recall from Sec.~\ref{sec:context-dependence} that the parser must represent spliced terms abstractly, by their location within the literal body. This implies that the TLM does not need to parse spliced expressions itself, as long as it can independently decide where the spliced expressions start and end. A paired start and end position is a segment,  represented by a value of the type \li{Segment.t} defined on Line 2 of Fig.~\ref{fig:relit-util}. Lines 26-27 and Lines 30-31 of the lexer produce a segment by calling a helper function, \li{Relit.Segment.read_to}, that internally invokes the Reason lexer on the provided lexing buffer until it sees an instance of the provided token outside of a Reason comment, string literal or generalized literal. If the provided token is a right delimiter also used by Reason, like \li{")"} on Line 4 of Fig.~\ref{fig:regex-lexer}, then \li{read_to} looks for the first unmatched instance. For example, if the remaining lexing buffer contains {\small \texttt{f("))"))A|G}} then \li{read_to} will consume up through the final instance of \li{)} and emit the start and end position, excluding the closing delimiter, relative to the original buffer.

This approach has two major benefits. First, it supports splicing even within literal notations that have a very different lexical structure from Reason itself, as in this example. Second, it allows the parser to avoiding needing to link to Reason's expression grammar (though this is also supported by Menhir). Spliced expressions arrive as single, opaque tokens as shown on Lines 6-7 of Fig.~\ref{fig:regex-parser}, and do not need to be parsed by the TLM at all. The system will recursively process spliced expressions after the parser has finished generating the proto-expansion.

Similar helper functions, not shown, are available for reading out single identifiers, simple paths like \li{X1.X2.xy} and a few other useful fragments of the Reason syntax, so that TLM providers can implement lower-cost splicing notation that clients can choose in situations where the spliced expression is of a restricted form that makes the end position of the spliced expression unambiguous without matching delimitation. As a matter of practice, we recommend that TLM providers start with a fully delimited form before adding optional alternative shorthand in order to avoid artificially restricting the client programmer's ability to directly convey the compositional structure of a value. 
Although \li{Segment.t} is not held abstract in order to support certain advanced scenarios, discussed below, we recommend that TLM providers generally treat it as such.

\subsubsection{\li{RegexParser} and Quasiquotation}\label{sec:regexparser}The job of a TLM parser is to generate a proto-expansion given the tokens generated by the lexer, or if this is not possible, to indicate an expansion error by raising either a Menhir parse error or \li{Relit.ExpansionError} (Line 5 of Fig.~\ref{fig:relit-util}) with an appropriate error message and, if possible, an error location. In the case of expression TLMs, the proto-expansion must be a proto-expression (we consider proto-patterns in the next section). The Menhir directive on Line 9 of Fig~\ref{fig:regex-parser}, which specifies the starting non-terminal of the grammar, reveals that proto-expressions are values of type \li{Relit.ProtoExpr.t}. This type is defined on Line 7 of Fig.~\ref{fig:relit-util} as a synonym for \li{Parsetree.expression}, which is the standard representation of expression parse trees exposed by the OCaml compiler \cite{ocaml-manual}.% by many external tools, including the Reason toolchain.

To be able to repurpose this existing parse tree representation for proto-expansions, there must be some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type. The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice references. Internally, we use OCaml's parse tree annotations.

>>>>

\begin{figure}
\begin{lstlisting}[deletekeywords={spliced}]
  module Segment : { 
    type t = {start_pos: int, end_pos: int};
    let read_to : string => Lexing.lexbuf => Segment.t; };
  exception ExpansionError({msg: string, loc: option(Segment.t)});
  module ProtoExpr : { 
    type t = Parsetree.expression; /* from the OCaml compiler library */
    let spliced : Segment.t => Parsetree.core_type => t; };
  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
  notation $proto_expr at ProtoExpr.t = $re_expr;
\end{lstlisting}
\vspace{-2px}
\caption{A fragment of the \li{Relit} module signature relevant to expression TLM providers.}
\label{fig:relit-util}
\vspace{-5px}
\end{figure}
    % let match_spliced : t => option(Segment.t, ProtoTyp.t);

\subsubsection{\li{RegexLexer}}
\label{sec:regex-lexer}
% A fragment of the signature of the \li{Relit_helper} library, which provides various necessary definitions and utilities to TLM providers, appears in Fig.~\ref{fig:relit-util}. 
In most respects, the definition of the lexer in Fig.~\ref{fig:regex-lexer} is conventional. It reads various token patterns on Lines 8-32, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}. Lines 22-23 combine sequences of non-special characters (e.g. alphanumeric characters) into a single string token. Lines 24-25 implement backslash-prefixed escape sequences for characters that have special meaning in this notation for regexes, emitting a string token in each instance.\footnote{A better approach would be to define a second lexing rule that greedily combines sequences consisting of non-special characters and escape characters into a single \li{STR} token, but for the sake of exposition, we stick to this simpler approach.}

Lines 24-31 of the lexer, which recognize the notation for regex splicing \li{$(}$e$\li{)}, and string splicing, \li{$$(}$e$\li{)}, are more unusual. Recall from Sec.~\ref{sec:context-dependence} that the parser must represent spliced terms abstractly, by their location within the literal body. This implies that the TLM does not need to parse spliced expressions itself, as long as it can independently decide where the spliced expressions start and end. A paired start and end position is a segment,  represented by a value of the type \li{Segment.t} defined in the \li{Relit} helper library as specified on Line 2 of Fig.~\ref{fig:relit-util}. Lines 26-28 and Lines 29-31 of the lexer produce a segment by calling a helper function, \li{Relit.Segment.read_to}, that internally invokes the Reason lexer on the provided lexing buffer until it sees an instance of the provided token outside of a Reason comment, string literal or generalized literal. If the provided token is a right delimiter also used by Reason, like \li{")"} on Line 4 of Fig.~\ref{fig:regex-lexer}, then \li{read_to} looks for the first unmatched instance. For example, if the remaining lexing buffer contains {\small \texttt{f("))"))A|G}} then \li{read_to} will consume up until just before the final instance of \li{)} and emit the appropriate segment. Similar helper functions, not shown, are available for reading out single identifiers, simple paths like \li{X1.X2.xy} and a few other unambiguously delimited fragments of the Reason syntax.

This approach has two major benefits. First, it supports splicing even within literal notations that have a very different lexical structure from Reason itself, as in this example. Second, it allows the parser to avoiding needing to link to Reason's expression grammar (though this is supported by Menhir), which substantially simplifies reasoning about ambiguities. Spliced expressions arrive as single, opaque tokens as specified on Lines 6-7 of Fig.~\ref{fig:regex-parser}. The system will recursively process spliced expressions after the parser has finished generating the proto-expansion.

%, so that TLM providers can implement lower-cost splicing notation that clients can choose in situations where the spliced expression is of a restricted form that makes the end position of the spliced expression unambiguous.

%As a matter of practice, we recommend that TLM providers start with a fully delimited form before adding optional alternative shorthand in order to avoid artificially restricting the client programmer's ability to directly convey the compositional structure of a value. 
% Although \li{Segment.t} is not held abstract in order to support certain advanced scenarios, discussed below, we recommend that TLM providers generally treat it as such.

\subsubsection{\li{RegexParser} and Quasiquotation}\label{sec:regexparser}The job of the parser is to generate a proto-expansion given the tokens generated by the lexer, or if this is not possible, to indicate an expansion error by raising either a Menhir parse error or \li{Relit.ExpansionError} (Line 4 of Fig.~\ref{fig:relit-util}) with an appropriate error message and, if possible, an error location. In the case of expression TLMs, the proto-expansion must be a proto-expression (we consider proto-patterns in the next section). Proto-expressions are encoded as values of type \li{Relit.ProtoExpr.t}. This type is defined on Line 6 of Fig.~\ref{fig:relit-util} as a synonym for \li{Parsetree.expression}, which is the standard representation of expression parse trees exposed by the OCaml compiler library \cite{ocaml-manual}.% by many external tools, including the Reason toolchain.

In order to repurpose this existing parse tree representation for proto-expansions, we must provide some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type. The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).\todo{say something about splice references in sec 6}{}
